A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-10 tex/sec4-pre-synthesis.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

\section{Mode Preprocessing and Related Properties}\label{sec:pre-synthesis}

Our first important construction is mode preprocessing in \cref{sec:mode-preprocessing}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
We also discuss some related properties:
By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
We also formulate \varcitet{Dunfield2021}{'s} annotatability and discuss its relationship with our completeness and generalised mode preprocessing in \cref{sec:annotatability}.

@@

\subsection{Generalised Mode Preprocessing}
\label{sec:mode-preprocessing}

The goal of this sub-section is to construct a mode preprocessor, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode preprocessor returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a preprocessor can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.%
\footnote{For clarity, we use ordinary mode preprocessing elsewhere in this paper.}

@@

Now we are ready to construct a generalised mode preprocessor.

\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
\end{theorem}

The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term and apply mode casts depending on what modes are required.
Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-preprocessing}.

@@

\begin{proof}[Proof of \cref{thm:generalised-mode-preprocessing}]
By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:strict-derivation}) by induction on~$t$.

@@

Finally, having constructed a generalised mode preprocessor, it is easy to derive an ordinary one.

\begin{corollary}[Mode Preprocessing]\label{thm:mode-preprocessing}
  It is decidable whether\/ $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
\end{corollary}

\begin{proof}
By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{proof}

%We can also establish the converse of \cref{thm:Pre?-true,thm:Pre?-false}.
%
%\begin{corollary}\label{thm:toPre?-true}
%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$ for some boolean~$s$.
%\end{corollary}
%
%\begin{corollary}\label{thm:toPre?-false}
%If\/ $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
%\end{corollary}
%
%Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
%But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
%And \cref{thm:toPre?-true} can also be proved easily in the same way.
%
%\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
%
%By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
%Symmetrically, if $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
%\end{proof}

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved from generalised mode preprocessing.

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode preprocessing does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.

>>>>

\section{Mode Decoration and Related Properties}\label{sec:pre-synthesis}

Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
We also discuss some related properties:
By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
We also formulate \varcitet{Dunfield2021}{'s} annotatability and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.

@@

\subsection{Generalised Mode Decoration}
\label{sec:mode-decoration}

The goal of this sub-section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
In fact we will do better:
If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
\footnote{For brevity, we use ordinary mode decoration elsewhere in this paper.}

@@

Now we are ready to construct a generalised mode decorator.

\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
\end{theorem}

The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term and apply mode casts depending on what modes are required.
Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-decoration}.

@@

\begin{proof}[Proof of \cref{thm:generalised-mode-decoration}]
By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:strict-derivation}) by induction on~$t$.

@@

Finally, having constructed a generalised mode decorator, it is easy to derive an ordinary one.

\begin{corollary}[Mode Preprocessing]\label{thm:mode-decoration}
  It is decidable whether\/ $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
\end{corollary}

\begin{proof}
By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{proof}

%We can also establish the converse of \cref{thm:Pre?-true,thm:Pre?-false}.
%
%\begin{corollary}\label{thm:toPre?-true}
%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$ for some boolean~$s$.
%\end{corollary}
%
%\begin{corollary}\label{thm:toPre?-false}
%If\/ $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
%\end{corollary}
%
%Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
%But this is exactly what generalised mode decoration does, so we can simply reuse \cref{thm:generalised-mode-decoration}.
%And \cref{thm:toPre?-true} can also be proved easily in the same way.
%
%\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
%
%By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
%Symmetrically, if $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
%\end{proof}

@@

Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.

### 2023-07-10 tex/sec7-future.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours: 
for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
Even for this mildly generalised setting, there is already a lot more work to do:
Both the mode preprocessor and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
There should be ways to fix this problem while retaining the mode preprocessing phase---for example, one idea is to make the mode preprocessor produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
All these considerations are rather too complex for a first theory, however, and should best be left as future work.

%\Josh[inline, caption={}]{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p
%\begin{enumerate}
%\item In general: two independently formulated type systems over raw terms
%\item More sensible next step: a `refinement chain'
%\item Backtracking algorithms are required. for both mode preprocessing and bidirectional type synthesis.
%\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.
%\item Too complex for a first theory, and left as future work
%\end{enumerate}}

>>>>

As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours: 
for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
Even for this mildly generalised setting, there is already a lot more work to do:
Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
All these considerations are rather too complex for a first theory, however, and we decide to leave them for future work.

%\Josh[inline, caption={}]{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p
%\begin{enumerate}
%\item In general: two independently formulated type systems over raw terms
%\item More sensible next step: a `refinement chain'
%\item Backtracking algorithms are required. for both mode decoration and bidirectional type synthesis.
%\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.
%\item Too complex for a first theory, and left as future work
%\end{enumerate}}
